Efficient pedigree recording for fast population genetics simulation

### 2018-01-21 forwards_paper.tex
Added Gil to the ACK list.
edits for AKs comments

\begin{figure}
    \begin{center}
        \includegraphics[width=\textwidth]{example_tree_sequence}
    \end{center}
    \caption{
        An example tree sequence with three samples over a chromosome of length 10.
        The left-hand panel shows the tree sequence pictorially in two different ways:
        (top) as a sequence of tree topologies
        and (bottom) the spatial extent of the edges that define these topologies.
        The right-hand panels show the specific encoding
        of this tree sequence in the four tables (nodes, edges, sites and mutations)
        defined by \msprime.
        \label{fig:example_tree_sequence}
    }
\end{figure}

@@

\begin{taocpalg}{W}{Forwards-time tree sequence}
{Simulates a randomly mating population of $N$ haploid individuals with
chromosome of length $L$ for $T$ generations, and returns the node
and edge tables ($\Nt$ and $\Et$) recording the simulated history.
The tables are simplified every $s$ generations, removing genealogical
information from $\Nt$ and $\Et$ irrelevant to the current population $P$.
}

@@

@@

\begin{figure}
    \begin{center}
        \includegraphics{method_diagram}
    \end{center}
    \caption{
        An example of tree sequence simplification.
        \textbf{(A)} The augmented pedigree diagram on the left
        relates the ten homologous chromosomes of five diploid individuals (BC, DE, FG, HI, and JK)
        to each other and to a common ancestral chromosome (A);
        dotted lines connect the two chromosomes of each individual,
        and solid lines lead to the products of their meioses.
        The corresponding tables (right) have 11 node records (one for each chromosome)
        and 15 edge records (one for each distinctly inherited segment).
        Blue numbers denote crossing over locations --
        for instance, $D$ and $E$ were parents to $G$,
        who inherited the left 70\% of the chromosome from $E$ and the remainder from $D$.
        $B$, $C$, $D$, and $E$ inherit clonally from $A$.
        \textbf{(B)} The five distinct trees
        found across the chromosome (blue numbers denote locations on the chromosome).
        Labels after simplification are shown in red.
        \textbf{(C)} Tables recording the tree sequence after simplification
        with nodes $J$ and $K$ as samples.
        The mapping from labels in the forwards time simulation to nodes in the tree sequence
        is shown in red.
        % which allows additional records to be added as the simulation progresses.
        \label{fig:method_diagram}
    }
\end{figure}

@@

Our approach to simplification is based on Hudson's algorithm for simulating
the coalescent with recombination~\citep{hudson1983properties},
paralleling the implementation in \msprime{} \citep{kelleher2016efficient};
an implementation in pseudocode is provided in Appendix~\ref{ss:simplify_algorithm},
and a python implementation as supplementary information.
Conceptually, this works by
(a) beginning by painting the chromosome in each sample a distinct color;
(b) moving back through history,
copying the colors of each chromosome to the portions of its' parental chromosomes
from which it was inherited;
(c) each time we would paint two colors in the same spot (a coalescence),
record that information as an edge and instead paint a brand-new color;
and
(d) once all colors have coalesced on a given segment,
stop propagating it.
This ``paint pot'' description misses some details --
for instance, we must ensure that all coalescing segments in a given individual
are assigned the \emph{same} new color --
but is reasonably close.
Figure~\ref{fig:method_diagram} shows an example tree sequence,
before and after simplification,
and Figure~\ref{fig:simplify_state} depicts the ``paint pot'' state of the algorithm
during the process of simplifying this tree sequence.

@@

%%%%%%%%%%%%%%%%%%%%%%
\section*{Acknowledgements}
Thanks to Jared Galloway, Brad Shaffer, and Evan McCartney--Melstad for useful discussions.
Work on this project was supported by funding from
the Sloan Foundation and the NSF (under DBI-1262645) to PR;
the Wellcome Trust (grant 100956/Z/13/Z) to Gil McVean;
the NIH (R01GM115564) to KRT.

>>>>

\begin{figure}
    \begin{center}
        \includegraphics[width=\textwidth]{example_tree_sequence}
    \end{center}
    \caption{
        An example tree sequence with three samples over a chromosome of length 10.
        The left-hand panels show the tree sequence pictorially in two different ways:
        (top) a sequence of tree topologies; the first tree extends from genomic position 0 to 5,
        and the second from 5 to 10; and
        (bottom) the edges that define these topologies,
        displayed over their corresponding genomic segment
        (for instance, the edge from node 2 to node 4 is present only on the interval from 0 to 5).
        The right-hand panels show the specific encoding
        of this tree sequence in the four tables (nodes, edges, sites and mutations)
        defined by \msprime.
        \label{fig:example_tree_sequence}
    }
\end{figure}

@@

\begin{taocpalg}{W}{Forwards-time tree sequence}
{Simulates a randomly mating population of $N$ haploid individuals with
chromosome of length $L$ for $T$ generations, and returns the node
and edge tables ($\Nt$ and $\Et$) recording the simulated history.
In each generation, the current population is stored in $P$,
while produced offspring are placed in $P'$.
The tables are simplified every $s$ generations, removing genealogical
information from $\Nt$ and $\Et$ irrelevant to the current population.
}

@@

\paragraph{Essential steps in tree sequence recording}
To record a tree sequence, including mutations,
a simulator must record:
\begin{itemize}
    \item the birth time of each individual in the Node Table,
    \item the genetic segments each individual inherits from her parents in the Edge Table, and
    \item the individual each new mutation occurs in, and the derived state in the Mutation Table,
        adding also to the Site Table if the site is previously unmutated.
\end{itemize}
Recall that here we mean ``\emph{haploid} individual'', i.e., gamete.
Before simplification, time must be translated to ``time ago'' if it is not already,
and the tables must be sorted.
Algorithm~\algref{W} avoided both of these details since $t$ denotes 
``time until the end of the simulation'',
and nonoverlapping generations ensure that the tables will be already sorted.
Finally, since $\tsimplify{P', \Nt, \Et}$ results in the $N$ individuals in $P'$
being relabeled in \msprime{}'s tables as $0, 1, \ldots, N-1$,
there must be bookkeeping that keeps in sync the individual IDs as recorded by the simulator
with the node IDs recorded in the tables.

@@

\begin{figure}
    \begin{center}
        \includegraphics{method_diagram}
    \end{center}
    \caption{
        An example of tree sequence simplification.
        \textbf{(A)} The augmented pedigree diagram on the left
        relates the ten homologous chromosomes of five diploid individuals (BC, DE, FG, HI, and JK)
        to each other and to a common ancestral chromosome (A);
        dotted lines connect the two chromosomes of each individual,
        and solid lines lead to the products of their meioses.
        The corresponding tables (right) have 11 node records (one for each chromosome)
        and 15 edge records (one for each distinctly inherited segment).
        Blue numbers denote crossing over locations --
        for instance, $D$ and $E$ were parents to $G$,
        who inherited the left 70\% of the chromosome from $E$ and the remainder from $D$.
        $B$, $C$, $D$, and $E$ inherit clonally from $A$.
        \textbf{(B)} The five distinct trees
        found across the chromosome (blue numbers denote locations on the chromosome).
        Labels after simplification are shown in red (see text).
        \textbf{(C)} Tables recording the tree sequence after simplification
        with nodes $J$ and $K$ as samples.
        The mapping from labels in the forwards time simulation to nodes in the tree sequence
        is shown in red.
        % which allows additional records to be added as the simulation progresses.
        \label{fig:method_diagram}
    }
\end{figure}

@@

Our approach to simplification is based on Hudson's algorithm for simulating
the coalescent with recombination~\citep{hudson1983properties},
paralleling the implementation in \msprime{} \citep{kelleher2016efficient};
an implementation in pseudocode is provided in Appendix~\ref{ss:simplify_algorithm},
and a python implementation as supplementary information.
Conceptually, this works by
(a) beginning by painting the chromosome in each sample a distinct color;
(b) moving back through history,
copying the colors of each chromosome to the portions of its' parental chromosomes
from which it was inherited;
(c) each time we would paint two colors in the same spot (a coalescence),
record that information as an edge and instead paint a brand-new color;
and
(d) once all colors have coalesced on a given segment,
stop propagating it.
This ``paint pot'' description misses some details --
for instance, we must ensure that all coalescing segments in a given individual
are assigned the \emph{same} new color --
but is reasonably close.
Figure~\ref{fig:method_diagram} shows an example tree sequence,
before and after simplification,
and Figure~\ref{fig:simplify_state} depicts the ``paint pot'' state of the algorithm
during the process of simplifying this tree sequence.
Since the method begins with the samples and moves back through time,
in the output tree sequence, the $n$ samples will be numbered $0, 1, \ldots, n-1$
and subsequent nodes will be ordered by time since birth.
This is seen in the red labels of Figure~\ref{fig:method_diagram}.

@@

%%%%%%%%%%%%%%%%%%%%%%
\section*{Acknowledgements}
Thanks to Gil McVean, Jared Galloway, Brad Shaffer, and Evan McCartney--Melstad for useful discussions.
Work on this project was supported by funding from
the Sloan Foundation and the NSF (under DBI-1262645) to PR;
the Wellcome Trust (grant 100956/Z/13/Z) to Gil McVean;
the NIH (R01GM115564) to KRT.
